PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2022-09-06 supplementary.tex
feat: continue supplementary

Once this link has been established we group simulations by the value of $h$ and check for accuracy comparing the reconstructed pangenome graphs with the ground truth provided by our simulation. In particular for each isolate we consider the breakpoints between different pancontigs that tile the genome, and measure the displacement between the real position of these breakpoints and the position reconstructed by pangraph.\\
To evaluate this displacement it is first necessary to link breakpoints on the real and reconstructed pangenome graph. We formulate this problem as an assignment problem, and solve it numerically using the Hungarian method \cite{kuhn1955hungarian}. For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of pangraph, if on any of the two pangenome graphs a pair of breakpoints sits at a distance smaller than this threshold, we remove one of them from the set of breakpoints.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of the isolate of pairs of breakpoints on the two graphs. We then numerically solve the assignment problem on this matrix, and obtain a set of $K = \min\{ I,J \}$ pairings between breakpoints such that the distance between the pairs is minimal and all the breakpoints of the graph with minimum number of breakpoints have been paired. The average distance between breakpoints in these pairs is the average breakpoint distance of the graph.\\
For each simulation we obtain a number of average breakpoint distances equal to the number of isolates in the simulation. In Fig.~\ref{fig:benchmark-accuracy-suppl} we plot the cumulative distribution of these distances, considering only distances smaller than 1 kbp, stratifying simulations according to the average pairwise divergence of pancontigs inferred from the value of $\mu$. Each panel correspond to a different alignment kernel. As the sequence diversity increases we observe a clear transition. For low diversity most of the breakpoints are inferred to be only a few bp away from their real position, while for highly diverged sequences the position of breakpoints is not precisely inferred and can be hundreds of bp away from their real position. For different alignment kernels this transition occurs at different values of the divergence, with \textit{mmseqs2} being accurate at higher divergence values, as can be visualize from Fig.~XX in the main text.

@@

We downloaded from RefSeq \cite{o2016reference} complete chromosomes from 5 different bacterial species, listed in Table~\ref{table:panx-dataset}.\footnote{A complete list of accession numbers can be found in...}. These genomes had previously been analyzed with PanX \cite{ding2018panx}. For each species we also report the average genome lenght and average pairwise divergence on core genome.

@@

\begin{table}[hb]
    \begin{tabular}{l l l l}
        \hline\hline
        Species         & N. isolates & Avg. chromosome length [Mbp] & Avg. pairwise divergence (core genes) \\
        \hline
        E.~Coli         & 307         & 4.98                         &                                       \\
        K.~Pneumoniae   & 109         & 5.35                         &                                       \\
        H.~Pylori       & 85          & 1.62                         &                                       \\
        M.~Tuberculosis & 51          & 4.40                         &                                       \\
        P.~Marinus      & 10          & 1.80                         &                                       \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset overview}. We downloaded complete chromosome sequences from RefSeq for 5 different bacterial species. For each species we report the number of isolates downloaded, along with the average chromosome length and average pairwise divergence on core genes.}
    \label{table:panx-dataset}
\end{table}

@@

We built pangenome graphs using five different options for the alignment kernel:
\begin{itemize}
    \item kernel \textit{minimap2} with option \textit{asm10} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{minimap2} with option \textit{asm20} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{mmseqs2} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{minimap2} with option \textit{asm10} and null energy parameters $\alpha=0$, $\beta=0$.
    \item kernel \textit{mmseqs2} and null energy parameters $\alpha=0$, $\beta=0$.
\end{itemize}
% TODO: describe expectations and reasons for the choice

@@

@@

Given a graph composed of many different genomes, PanGraph is capable of marginalizing this graph on a subset of strains with the \verb|marginalize| command. To test this feature we built pangenome graphs on genomes from 5 different species: \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC), extracted from RefSeq.\footnote{The list of selected accession numbers for each species and pair is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each strain we took 50 different strains (10 for PM) and built an unique pangenome graph using the \verb|build| command with standard values of the parameters. For each species we then randomly picked 50 different pairs of strains (45 for PM) and extracted the corresponding sub-graphs using the \verb|marginalize| command. We compared each of these graphs to their counterpart generated simply by using the \verb|build| command on the corresponding pair of strains.\\
The comparison between the marginalized (MG) and pairwise (PW) graph consists in splitting each genome using the overlap of the breakpoints given by the two different graphs. Then for each resulting segment we check whether it is shared with the other strain of the pair on either of the graphs. This gives raise to four different possibilities, corresponding to the segment being shared (1) or not (0) on either of the two graphs (MG/PW). We consider four different classes of segments: \textit{agree on sharing} (1-1 or 0-0), \textit{shared on both} (1-1), \textit{private on both} (0-0) and \textit{disagree on sharing} (1-0 or 0-1). For each class we measure the total and average length of the segments belonging to that class, and display the results in Fig.~\ref{fig:marginalize-suppl}.\\
For all species except for PM we find that MG graphs are compatible with their PW counterpart: segments on which the graphs \textit{disagree on sharing} represent only a minor fraction of the genome (much smaller than the fraction of shared or private segments) and their size is compatible with the default block length threshold of PanGraph (100 bp). PM represents and exception, with the great majority of segments being private on each graph and having large size. This expected since for PM the strains considered are highly diverged, above the sensitivity of minimap2 ($\sim 27\%$ core divergence). This causes the majority of blocks not to be merged, and generates a strong order dependence. The size of segments with \textit{disagreement on sharing} is also the largest, some segments having size of 10s of kbp.

>>>>

Once this link has been established we group simulations by the value of $h$ and check for accuracy comparing the reconstructed pangenome graphs with the ground truth provided by our simulation. In particular for each isolate we consider the breakpoints between different pancontigs that tile the genome, and measure the displacement between the real position of these breakpoints and the position reconstructed by pangraph.\\
To evaluate this displacement it is first necessary to link breakpoints on the real and reconstructed pangenome graph. We formulate this problem as an assignment problem, and solve it numerically using the Hungarian method \cite{kuhn1955hungarian}. For each isolate we call $b_i$ for $i=1,\ldots,I$ the panconting breakpoints on the real pangenome graph, and $b_j$ for $j=1,\ldots,J$ the ones on the reconstructed pangenome graph.\footnote{To avoid artifacts generated by the default threshold distance $d=100$ bp of pangraph, if on any of the two pangenome graphs a pair of breakpoints sits at a distance smaller than this threshold, we remove one of them from the set of breakpoints.} We define a cost matrix $D_{ij} = d(b_i,b_j)$ whose elements are distances on the genome of pairs of breakpoints from the two graphs. We then numerically solve the assignment problem on this matrix, and obtain a set of $K = \min\{ I,J \}$ pairings between breakpoints such that the distance between the pairs is minimal and all the breakpoints of the graph with minimum number of breakpoints have been paired. The average distance between breakpoints in these pairs is the average breakpoint distance of the graph.\\
For each simulation we obtain a number of average breakpoint distances equal to the number of isolates in the simulation. In Fig.~\ref{fig:benchmark-accuracy-suppl} we plot the cumulative distribution of these distances, considering only distances smaller than 1 kbp, stratifying simulations according to the average pairwise divergence of pancontigs inferred from the value of $\mu$. Each panel correspond to a different alignment kernel. As the sequence diversity increases we observe a clear transition. For low diversity most of the breakpoints are inferred to be only a few bp away from their real position, while for highly diverged sequences the position of breakpoints is not precisely inferred and can be hundreds of bp away from their real position. For different alignment kernels this transition occurs at different values of the divergence, with \textit{mmseqs2} being accurate at higher divergence values, as can be visualize from Fig.~XX in the main text.

@@

We downloaded from RefSeq \cite{o2016reference} complete chromosomes from 5 different bacterial species:\footnote{A complete list of accession numbers can be found in...} \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC). The number of isolates selected per species, average chromosome length and average pairwise divergence on core genes are reported in Table~\ref{table:panx-dataset}. These genomes had previously been analyzed with PanX \cite{ding2018panx}.

@@

\begin{table}[hb]
    \begin{tabular}{l l l l}
        \hline\hline
        Species         & N. isolates & Avg. chromosome length [Mbp] & Avg. pairwise divergence (core genes) \\
        \hline
        E.~Coli         & 307         & 5.0                          &                                       \\
        K.~Pneumoniae   & 109         & 5.3                          &                                       \\
        H.~Pylori       & 85          & 1.6                          &                                       \\
        M.~Tuberculosis & 51          & 4.4                          &                                       \\
        P.~Marinus      & 10          & 1.8                          &                                       \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset overview}. We downloaded complete chromosome sequences from RefSeq for 5 different bacterial species. For each species we report the number of isolates downloaded, along with the average chromosome length and average pairwise divergence on core genes.}
    \label{table:panx-dataset}
\end{table}

@@

We built pangenome graphs using five different options for the alignment kernel:
\begin{itemize}
    \item kernel \textit{minimap2} with option \textit{asm10} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{minimap2} with option \textit{asm20} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{mmseqs2} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{minimap2} with option \textit{asm10} and null energy parameters $\alpha=0$, $\beta=0$.
    \item kernel \textit{mmseqs2} and null energy parameters $\alpha=0$, $\beta=0$.
\end{itemize}
From the results described in the previous section we expect \textit{mmseqs2} to be able to merge sequences with higher divergence than \textit{minimap2}. Moreover from the definition of the pseudo-energy \textbf{eq.~1} in the main text, the value of the energy parameters $\beta$ defines an upper threshold for sequence divergence $d \sim 1/\beta$. Homologous sequences with divergence higher than this value will result in positive pseudo-energy and will not be merged in a single pancontig. For $\beta=10$ the critical divergence threshold is $d \sim 10\%$. Executing pangraph with the option $\alpha=0$, $\beta=0$ will result in the merging of more diverged sequences, at the cost of a more fragmented pangenome graph (more but shorter pancontigs).

@@

In Fig.~\ref{fig:panx-benchmark-suppl} we display the results of the benchmakr. On every panel we display on the x-axis results for different species, and indicate with different colors the five possible choices for the alignment kernel.

@@

% TODO: simplify dataset descritpion (is already described above)

Given a graph composed of many different genomes, PanGraph is capable of marginalizing this graph on a subset of strains with the \verb|marginalize| command. To test this feature we built pangenome graphs on a subset of genomes from the 5 different species we previously analyzed. For each species we took 50 different strains (10 for PM)\footnote{The list of selected accession numbers for each species and pair is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} and built an unique pangenome graph using the \verb|build| command with standard values of the parameters. For each species we then randomly picked 50 different pairs of strains (45 for PM) and extracted the corresponding sub-graphs using the \verb|marginalize| command. We compared each of these graphs to their counterpart generated simply by using the \verb|build| command on the corresponding pair of strains.\\
The comparison between the marginalized (MG) and pairwise (PW) graph consists in splitting each genome using the overlap of the breakpoints given by the two different graphs. Then for each resulting segment we check whether it is shared with the other strain of the pair on either of the graphs. This gives raise to four different possibilities, corresponding to the segment being shared (1) or not (0) on either of the two graphs (MG/PW). We consider four different classes of segments: \textit{agree on sharing} (1-1 or 0-0), \textit{shared on both} (1-1), \textit{private on both} (0-0) and \textit{disagree on sharing} (1-0 or 0-1). For each class we measure the total and average length of the segments belonging to that class, and display the results in Fig.~\ref{fig:marginalize-suppl}.\\
For all species except for PM we find that MG graphs are compatible with their PW counterpart: segments on which the graphs \textit{disagree on sharing} represent only a minor fraction of the genome (much smaller than the fraction of shared or private segments) and their size is compatible with the default block length threshold of PanGraph (100 bp). PM represents and exception, with the great majority of segments being private on each graph and having large size. This expected since for PM the strains considered are highly diverged, above the sensitivity of minimap2 ($\sim 27\%$ core divergence). This causes the majority of blocks not to be merged, and generates a strong order dependence. The size of segments with \textit{disagreement on sharing} is also the largest, some segments having size of 10s of kbp.
